Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-02-19 index.qmd
Minor updates

Slides: [slides.html](slides.html){target="_blank"}

Literature Comments: [Literature.html](articleSummaries.html)

@@

As the COVID-19 pandemic demonstrated, the ability to predict the
survival of a patient based on their symptoms and medical history is a
major benefit as society deals with health crises such as pandemics.
Those predictions would assist medical practitioners in triage and allow
researchers the ability to generate predictions on the individuals who
are at the greatest risk of losing their lives due to an illness. While
the medical world is the most obvious use of survival predictions, the
same methods can be applied to the predicting bank failure, the
remaining useful life of machines, machinery maintenance timelines, and
insurance likelihood of payouts. There are various survival methods that
can be examined, including Cox proportional hazards model which will be
examined in this paper.

@@

An additional limitation of CPH comes from the data in the covariates
selected and how can they can become biased. Two studies done using UK
health data demonstrated how bias can impact the values for the
covariates when relying on patient submitted information. The first was
a study on the connection between salt intake and the development of
anxiety or depression after 14.5 years [@wang2025] and the second was a
study on the association between postpartum depression and the
development of two or more chronic diseases [@zhang2025]. Both of those
studies relied on patient submitted health data and survey responses
that were supplemented with death records and other accessible health
records thus creating a possible bias in their data which could
negatively influence their model as their covariates may not have
accurately reflected the true data. These articles provide examples of
the care that needs to be taken when working with CPH models.

While health care is the dominate field in which CPH is used, a study in
2021 provides an example of using CPH in insurance fields. An insurance
company in the Czech Republic provides policies for certain illness
events until the event occurs when the insured will receive money and
their contract is complete. To assist the company with predicting if an
individual will have one of the illness events occur, a model using CPH
was generated using gender, age, and region as the
covariates[@zapletal2021]. A study in 2023 focused on the length of
productive life of female Floridian goats to determine the factors that
would impact the longevity of the goats. CPH was used to determine that
the age of first kidding and the specific herd impacts the length of
productive life of the goats [@ziadi2023]. Both of these studies provide
examples of the reaches of CPH models outside the health industry.

Continuing with using CPH outside of health data, we will study the
remaining useful life of a NASA aircraft engine using data provided by
NASA for the study on propagation modeling [@saxena2008]. From the
literature review, we have noted that we will need to be concerned with
the constant hazards ratio within our data and determine how we will
accommodate censored data when creating our model.

<!-- -->

@@

@@

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total))

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()

### Modeling and Results

>>>>

[Slides](slides.html){target="_blank"} \|
[Literature](articleSummaries.html)

@@

As the COVID-19 pandemic demonstrated, the ability to predict the
survival of a patient based on their symptoms and medical history is a
major benefit as society deals with health crises such as pandemics.
Those predictions would assist medical practitioners in triage and allow
researchers the ability to generate predictions on the individuals who
are at the greatest risk of losing their lives due to an illness. While
the medical world is the most obvious use of survival predictions, the
same methods can be applied to predicting bank failure, the remaining
useful life of machines, machinery maintenance timelines, and insurance
likelihood of payouts. There are various survival methods that can be
examined, including Cox proportional hazards model which will be
examined in this paper.

@@

An additional limitation of CPH comes from the data in the covariates
selected and how they can become biased. Two studies done using UK
health data demonstrated how bias can impact the values for the
covariates when relying on patient submitted information. The first was
a study on the connection between salt intake and the development of
anxiety or depression after 14.5 years [@wang2025] and the second was a
study on the association between postpartum depression and the
development of two or more chronic diseases [@zhang2025]. Both of those
studies relied on patient submitted health data and survey responses
that were supplemented with death records and other accessible health
records thus creating a possible bias in their data which could
negatively influence their model as their covariates may not have
accurately reflected the true data. These articles provide examples of
the care that needs to be taken when working with CPH models.

While health care is the dominant field in which CPH is used, a study in
2021 provides an example of using CPH in insurance fields. An insurance
company in the Czech Republic provides policies for certain illness
events until the event occurs when the insured will receive money and
their contract is complete. To assist the company with predicting if an
individual will have one of the illness events occur, a model using CPH
was generated using gender, age, and region as the
covariates[@zapletal2021]. A study in 2023 focused on the length of
productive life of female Floridian goats to determine the factors that
would impact the longevity of the goats. CPH was used to determine that
the age of first kidding and the specific herd impacts the length of
productive life of the goats [@ziadi2023]. Both of these studies provide
examples of the reaches of CPH models outside the health industry.

CPH has the capability to be applied to many fields, though it poses
some limitations. The constant hazards ratio and determining how to
censor data are two concerns that can lead to limitations in using CPH
for models. In addition, the necessity of a time to event value within
the data source is something that needs to be taken into consideration
when choosing to use CPH. Without that value, using CPH as a survival
model will not be the appropriate method to take with that particular
data. This value is also what makes CPH a prime modeling candidate with
health data as health metrics frequently have a time component to them,
especially when comparing someone over a period of time in a health
study as is done with medications and the development of cancer or other
illnesses. However, while the usage of CPH is clear in the medical
field, this paper will demonstrate using CPH within the field of
mechanics and remaining useful life of a machine.

@@

Though CPH is typically used with health data, we wanted to determine
how CPH can be used in other fields. As such, we will study the
remaining useful life of a NASA aircraft engine using data provided by
NASA for the study on propagation modeling [@saxena2008]. In particular,
we will be examining their data from what they have deemed engine two.
From the literature review, we have noted that we will need to be
concerned with the constant hazards ratio within our data and determine
how we will accommodate censored data when creating our model.

We will be using the test_FD002 and train_FD002 datasets available for
download at [NASA's Open Data
Portal](https://data.nasa.gov/Aerospace/CMAPSS-Jet-Engine-Simulated-Data/ff5v-kuh6/about_data)
and also available on the home page of our [GitHub
repository](https://github.com/JR-87/capstone4ds_reed).

```{r}
testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'  
test <- read.table(testurl, header = TRUE, sep = '')
head(test)
```

```{r}
trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'  
test <- read.table(trainurl, header = TRUE, sep = '') 
head(test)
```

##

@@

A study was conducted to determine how...Modeling and Results
